Dependency in Refinement Logics

### 2015-10-18 doc/paper.tex
Fix missing subscript
demonstrate rules for dependent pair, function

@@

\def\InputModeColorName{MidnightBlue}
\def\OutputModeColorName{Maroon}
\newcommand\InputMode[1]{{\color{\InputModeColorName}{#1}}}
\newcommand\OutputMode[1]{{\color{\OutputModeColorName}{#1}}}
\newcommand\Quote[1]{{\color{Gray}{«}}#1{\color{Gray}{»}}}

\newcommand\HypJ[2]{#1\ \ (#2)}
\newcommand\GenJ[2]{\vert_{\InputMode{#1}}\; #2}
\newcommand\JJ{J}
\newcommand\Tuple[1]{\left\langle#1\right\rangle}
\newcommand\HTuple[1]{\left\lBrace#1\right\rBrace}
\newcommand\RSet{\mathfrak{R}}
\newcommand\JSet{\mathfrak{J}}

\newcommand\IsTac[1]{\InputMode{#1}\ \textit{tactic}}

@@

%% subgoals
\newcommand\ISG[2]{\InputMode{#1}\,{\color{Black}:}\,\InputMode{#2}}
\newcommand\OSG[2]{\InputMode{#1}\,{\color{Black}:}\,\OutputMode{#2}}

%% refinement
\newcommand\Refine[4]{\InputMode{#1}\Vdash\InputMode{#2}\Rightarrow\OutputMode{#4}\dashv\OutputMode{#3}}
\newcommand\NoRefine[2]{\InputMode{#1}\Vdash\InputMode{#2}\Uparrow}

\newcommand\ThenMultiAux[5]{%
  \InputMode{#2}\dashv\InputMode{#3}%
  \Vdash^\star\InputMode{#1}\Rightarrow%
  \OutputMode{#4}\dashv\OutputMode{#5}%
}

\newcommand\Member[2]{\InputMode{#1}\in\InputMode{#2}}
\newcommand\MemberUnmoded[2]{#1\in #2}
\newcommand\IsTrue[1]{\InputMode{#1}\ \textit{true}}
\newcommand\IsTrueUnmoded[1]{#1\ \textit{true}}
\newcommand\IdTac{\mathtt{id}}
\newcommand\FailTac{\mathtt{fail}}
\newcommand\OrElseTac[2]{#1\vert#2}
\newcommand\ThenTac[2]{#1;#2}
\newcommand\ThenLTac[2]{#1;[#2]}
\newcommand\DProdIntroRule[1]{\mathtt{PairIntro}\{#1\}}
\newcommand\DProdIntroRuleNullary{\mathtt{PairIntro}}
\newcommand\TyProd[2]{#1 \times #2}
\newcommand\TyDProd[3]{(#2:#1) \times #3}
\newcommand\TyMember[2]{#1\in#2}
\newcommand\Dom[1]{\left\vert#1\right\vert}
\newcommand\SubsetEq[2]{\InputMode{#1}\subseteq\InputMode{#2}}
\newcommand\FV[1]{\textit{FV}\left(#1\right)}
\newcommand\Subst[3]{\left[#1\;/\;#2\right]#3}
\newcommand\ModLCF{\textbf{\emph{Modernized LCF}}}
\newcommand\DepLCF{\textbf{\emph{Dependent LCF}}}

\newcommand\IsEquiv[2]{\InputMode{#1}\equiv\InputMode{#2}}

@@

The sequencing tactical
$\ThenTac{\tau_1}{\MkFam{\tau}{\alpha}{I}}$ corresponds to
\verb!THENL! in LCF; the uniform sequencing tactical \verb!THEN! may be
recovered as a definitional extension, with its second argument given as the
constant family:
\[
    \InputMode{\ThenTac{\tau_1}{\tau_2}}
    \triangleq
    \OutputMode{\ThenTac{\tau_1}{\MkSet{\tau_2}}}
\]

@@

An LCF-style refiner has a property called \emph{constructible subgoals}, which
means that the subgoals incurred by a rule or tactic may be constructed
independently, using only the statement of the main goal. One unfortunate
consequence of this principle is that it is impossible to define a refinement
rule which expresses a dependency between subgoals. The canonical example is
the introduction rule for dependent pairs in Type Theory:
\[
  \infer[\textit{PairIntro}]{
    \Member{\Tuple{M,N}}{\TyDProd{A}{x}{B}}
  }{
    \Member{M}{A} &
    \Member{N}{\Subst{M}{x}{B}}
  }
\]

However, consider what a refinement rule for this would look like; we would
like to take the goal $\IsTrue{\TyDProd{A}{x}{B}}$ to two subgoals (one for
each conjunct); the first subgoal $\IsTrue{A}$ is clear enough, but it is not
possible to even write down the second subgoal until we know the synthesis of
the first one.

Because a tactic in \ModLCF\ produces only a \emph{context} of independent
subgoals, we cannot give a refinement treatment to this rule, and must
instead write a family of refinement rules $\DProdIntroRule{w}$ fibred over
witnesses of the left conjunct $w$:

\[
  \Refine{\IsTrueUnmoded{\TyDProd{A}{x}{B}}}{\DProdIntroRule{w}}{%
    \left\{\begin{array}{l}
      \OSG{\alpha}{\MemberUnmoded{w}{A}}\\
      \OSG{\beta}{\IsTrueUnmoded{\Subst{w}{x}{B}}}
    \end{array}\right.
  }{\Tuple{w,\beta}}
\]

@@

We can now encode \emph{PairIntro} with a single refinement rule,
$\DProdIntroRuleNullary$:

\[
  \Refine{\IsTrueUnmoded{\TyDProd{A}{x}{B}}}{\DProdIntroRuleNullary}{%
    \left\{\begin{array}{l}
        \OSG{\alpha}{\IsTrueUnmoded{A}}\\
        \OSG{\beta}{\IsTrueUnmoded{\Subst{\alpha}{x}{B}}}
    \end{array}\right.
  }{
    \Tuple{\alpha,\beta}
  }
\]

@@

\begin{gather*}
  \infer{
    \ThenMultiAux{\MkSet{}}{E}{\cdot}{E}{\cdot}
  }{
  }\\[6pt]
  %
  \infer{
    \ThenMultiAux{
      \MkFam{\tau}{\beta}{\Dom{\ISG{\alpha}{\JJ},\Phi}}
    }{E}{
      \ISG{\alpha}{\JJ},\Phi
    }{E'}{
      \Psi_\alpha\oplus\Phi'
    }
  }{
    \Refine{\JJ}{\tau}{\Psi_\alpha}{E_\alpha} &
    \ThenMultiAux{
      \MkFam{\tau}{\beta}{\Phi}
    }{
      \Subst{E_\alpha}{\alpha}{E}%
    }{
      \Subst{E_\alpha}{\alpha}{\Phi}
    }{E'}{\Phi'}
  }
\end{gather*}

@@

%% \section{ML Implementation of \DepLCF}

>>>>

\definecolor{dark-gray}{gray}{0.20}

@@

\def\IModeColorName{MidnightBlue}
\def\OModeColorName{Maroon}
\newcommand\IMode[1]{{\color{\IModeColorName}{#1}}}
\newcommand\OMode[1]{{\color{\OModeColorName}{#1}}}
\newcommand\Quote[1]{{\color{Gray}{«}}#1{\color{Gray}{»}}}

\newcommand\HypJ[2]{#1\ \ (#2)}
\newcommand\GenJ[2]{\vert_{\IMode{#1}}\; #2}
\newcommand\JJ{J}
\newcommand\RSet{\mathfrak{R}}
\newcommand\JSet{\mathfrak{J}}

\newcommand\IsTac[1]{\IMode{#1}\ \textit{tactic}}

@@

%% subgoals
\newcommand\ISG[2]{\IMode{#1}\,{\color{Black}:}\,\IMode{#2}}
\newcommand\OSG[2]{\IMode{#1}\,{\color{Black}:}\,\OMode{#2}}

%% refinement
\newcommand\Refine[4]{\IMode{#1}\Vdash\IMode{#2}\Rightarrow\OMode{#4}\dashv\OMode{#3}}
\newcommand\NoRefine[2]{\IMode{#1}\Vdash\IMode{#2}\Uparrow}

\newcommand\ThenMultiAux[5]{%
  \IMode{#2}\dashv\IMode{#3}%
  \Vdash^\star\IMode{#1}\Rightarrow%
  \OMode{#4}\dashv\OMode{#5}%
}

\newcommand\Member[2]{\IMode{#1}\in\IMode{#2}}
\newcommand\MemberUnmoded[2]{#1\in #2}
\newcommand\EqMemberUnmoded[3]{#1 = #2\in #3}
\newcommand\IsTrue[1]{\IMode{#1}\ \textit{true}}
\newcommand\IsTrueUnmoded[1]{#1\ \textit{true}}
\newcommand\IdTac{\mathtt{id}}
\newcommand\FailTac{\mathtt{fail}}
\newcommand\OrElseTac[2]{#1\vert#2}
\newcommand\ThenTac[2]{#1;#2}
\newcommand\ThenLTac[2]{#1;[#2]}
\newcommand\DPairIntroRule[1]{\mathtt{PairIntro}\{#1\}}
\newcommand\DPairIntroRuleNullary{\mathtt{PairIntro}}
\newcommand\DPairElimRule[1]{\mathtt{PairElim}\{#1\}}
\newcommand\DFunIntroRule{\mathtt{FunIntro}}
\newcommand\DFunElimRule[1]{\mathtt{FunElim}\{#1\}}
\newcommand\TyDPair[3]{(#2:#1) \times #3}
\newcommand\TyDFun[3]{(#2:#1) \to #3}
\newcommand\Lam[2]{\lambda(#1.#2)}
\newcommand\Ap[2]{\mathsf{ap}(#1;#2)}
\newcommand\Ax{\mathsf{Ax}}
\newcommand\Tuple[1]{\left\langle#1\right\rangle}
\newcommand\Spread[4]{\mathsf{spread}(#1;#2,#3.#4)}
\newcommand\TyUniv{\mathbb{U}}
\newcommand\TyMember[2]{#1\in#2}
\newcommand\Dom[1]{\left\vert#1\right\vert}
\newcommand\SubsetEq[2]{\IMode{#1}\subseteq\IMode{#2}}
\newcommand\FV[1]{\textit{FV}\left(#1\right)}
\newcommand\Subst[3]{\left[#1\;/\;#2\right]#3}
\newcommand\ModLCF{\textbf{\emph{Modernized LCF}}}
\newcommand\DepLCF{\textbf{\emph{Dependent LCF}}}

\newcommand\IsEquiv[2]{\IMode{#1}\equiv\IMode{#2}}

\newcommand\Seq[2]{#1 \gg #2}

\newcommand\Ext[1]{{\color{dark-gray}\left\lfloor\mathbf{ext}\ \OMode{#1}\right\rfloor}}
\newcommand\SeqExt[3]{
  \Seq{\IMode{#1}}{\IMode{#2}}\ \Ext{#3}
}
\newcommand\NuprlRule[3]{%
  \begin{array}{l}%
    #2\\
    \ \ \mathbf{by}\ #1\\
    \quad #3
  \end{array}%
}

@@

The sequencing tactical
$\ThenTac{\tau_1}{\MkFam{\tau}{\alpha}{I}}$ corresponds to
\verb!THENL! in LCF; the uniform sequencing tactical \verb!THEN! may be
recovered as a definitional extension, with its second argument given as the
constant family:
\[
    \IMode{\ThenTac{\tau_1}{\tau_2}}
    \triangleq
    \OMode{\ThenTac{\tau_1}{\MkSet{\tau_2}}}
\]

@@

An LCF-style refiner has a property called \emph{constructible subgoals}, which
means that the subgoals incurred by a rule or tactic may be constructed
independently, using only the statement of the main goal. One unfortunate
consequence of this principle is that it is impossible to define a refinement
rule which expresses a dependency between subgoals. The canonical example is
the introduction rule for dependent pairs in Type Theory:
\[
  \infer[\textit{PairIntro}]{
    \Member{\Tuple{M,N}}{\TyDPair{A}{x}{B}}
  }{
    \Member{M}{A} &
    \Member{N}{\Subst{M}{x}{B}}
  }
\]

However, consider what a refinement rule for this would look like; we would
like to take the goal $\IsTrue{\TyDPair{A}{x}{B}}$ to two subgoals (one for
each conjunct); the first subgoal $\IsTrue{A}$ is clear enough, but it is not
possible to even write down the second subgoal until we know the synthesis of
the first one.

Because a tactic in \ModLCF\ produces only a \emph{context} of independent
subgoals, we cannot give a refinement treatment to this rule, and must
instead write a family of refinement rules $\DPairIntroRule{w}$ fibred over
witnesses of the left conjunct $w$:

\[
  \Refine{\IsTrueUnmoded{\TyDPair{A}{x}{B}}}{\DPairIntroRule{w}}{%
    \left\{\begin{array}{l}
      \OSG{\alpha}{\MemberUnmoded{w}{A}}\\
      \OSG{\beta}{\IsTrueUnmoded{\Subst{w}{x}{B}}}
    \end{array}\right.
  }{\Tuple{w,\beta}}
\]

@@

We can now encode \emph{PairIntro} with a single refinement rule,
$\DPairIntroRuleNullary$:

\[
  \Refine{\IsTrueUnmoded{\TyDPair{A}{x}{B}}}{\DPairIntroRuleNullary}{%
    \left\{\begin{array}{l}
        \OSG{\alpha}{\IsTrueUnmoded{A}}\\
        \OSG{\beta}{\IsTrueUnmoded{\Subst{\alpha}{x}{B}}}
    \end{array}\right.
  }{
    \Tuple{\alpha,\beta}
  }
\]

@@

\begin{gather*}
  \infer{
    \ThenMultiAux{\MkSet{}}{E}{\cdot}{E}{\cdot}
  }{
  }\\[6pt]
  %
  \infer{
    \ThenMultiAux{
      \MkFam{\tau}{\beta}{\MkSet{\alpha}\cup\Dom\Phi}
    }{E}{
      \ISG{\alpha}{\JJ},\Phi
    }{E'}{
      \Psi_\alpha\oplus\Phi'
    }
  }{
    \Refine{\JJ}{\tau_\alpha}{\Psi_\alpha}{E_\alpha} &
    \ThenMultiAux{
      \MkFam{\tau}{\beta}{\Phi}
    }{
      \Subst{E_\alpha}{\alpha}{E}%
    }{
      \Subst{E_\alpha}{\alpha}{\Phi}
    }{E'}{\Phi'}
  }
\end{gather*}

@@

\section{Case Study: Constructive Type Theory}

The refinement logic for CTT is a sequent calculus with a single form of
judgment, \framebox{$\Seq{\IMode{H}}{\IMode{P}}$}, which means that the
proposition $P$ is true (and functionally so) under the assumptions in the
telescope $H$. The synthesis of this judgment is a hypothetical witness for the
truth or $P$ (or \emph{extract term}) with free variables in $H$.

The type theory CTT is then approximated in the refinement logic by adding
valid refinement rules for $\Seq{\IMode{H}}{\IMode{P}}$. For the sake of
familiarity, we will re-use Nuprl's rule notation such that
\[
  \NuprlRule{\texttt{RuleName}}{
    \SeqExt{H}{P}{E(\alpha_0,\alpha_1,\dots,\alpha_n)}
  }{
    \begin{array}{l}
      \SeqExt{H_0}{Q_0}{\alpha_0}\\
      \SeqExt{H_1(\alpha_0)}{Q_1(\alpha_0)}{\alpha_1(\alpha_0)}\\
      \vdots\\
      \SeqExt{H_n(\alpha_0,\alpha_1,\dots,\alpha_{n-1})}{Q_n(\alpha_0,\alpha_1,\dots,\alpha_{n-1})}{\alpha_n(\alpha_0,\dots,\alpha_{n-1})}
    \end{array}
  }
\]
shall mean
\[
  \Refine{
    \Seq{H}{P}
  }{
    \mathtt{RuleName}
  }{
    \left\{\begin{array}{l}
      \OSG{\alpha_0}{\Seq{H_0}{Q_0}}\\
      \OSG{\alpha_1}{\Seq{H_1}{Q_1}}\\
      \multicolumn{1}{c}{\color{Black}\vdots}\\
      \OSG{\alpha_n}{\Seq{H_n}{Q_n}}\\
    \end{array}\right.
  }{
    E(\alpha_0,\alpha_1,\dots,\alpha_n)
  }
\]

Let us first give some of the refinement rules for the dependent pair type:

\begin{gather*}
  \NuprlRule{\DPairIntroRuleNullary}{
    \SeqExt{H}{\TyDPair{A}{x}{B}}{\Tuple{\alpha,\beta}}
  }{
    \begin{array}{l}
      \SeqExt{H}{A}{\alpha}\\
      \SeqExt{H}{\Subst{\alpha}{x}{B}}{\beta}\\
      \SeqExt{H,x:A}{\MemberUnmoded{B}{\TyUniv}}{\gamma(x)}
    \end{array}
  }\\[6pt]
  \NuprlRule{\DPairElimRule{z}}{
    \SeqExt{H,z:\TyDPair{A}{x}{B},J}{C}{
      \Spread{z}{u}{v}{\alpha(u,v)}
    }
  }{
    \SeqExt{H,z:\TyDPair{A}{x}{B},u:A,v:\Subst{u}{x}{B},\Subst{\Tuple{u,v}}{z}{J}}{C}{
      \alpha(u,v)
    }
  }
\end{gather*}

We can also give refinement rules for dependent function introduction and
elimination; in the same way \ModLCF\ could not express a refinement rule for
pair introduction, a refinement rule for function elimination likewise only
becomes attainable via the generalization to \DepLCF.

\begin{gather*}
  \NuprlRule{\DFunIntroRule}{
    \SeqExt{H}{\TyDFun{A}{x}{B}}{\Lam{x}{\alpha(x)}}
  }{
    \begin{array}{l}
      \SeqExt{H,x:A}{B}{\alpha(x)}\\
      \SeqExt{H}{\MemberUnmoded{A}{\TyUniv}}{\beta}
    \end{array}
  }\\[6pt]
  \NuprlRule{\DFunElimRule{z}}{
    \SeqExt{
      H,z:\TyDFun{A}{x}{B},J
    }{C}{
      \beta(\Ap{z}{\alpha}, \Ax)
    }
  }{
    \begin{array}{l}
      \SeqExt{H,z:\TyDFun{A}{x}{B}, J}{A}{\alpha}\\
      \SeqExt{
        H,z:\TyDFun{A}{x}{B},%
        y:\Subst{\alpha}{x}{B},%
        p:\EqMemberUnmoded{y}{\Ap{z}{\alpha}}{\Subst{\alpha}{x}{B}}, J
      }{C}{
        \beta(y,p)
      }
    \end{array}
  }
\end{gather*}
